---
title: "bellabeat case study"
author: "hassan"
date: "2023-06-10"
output: html_document
---


# Data prepare 

```{r}
 setwd("~/bella")
library(tidyverse)
library(ggplot2)
library(lubridate)

```

```{r}
daily_activity <- read.csv("dailyActivity_merged.csv")
sleepday <- read.csv("sleepDay_merged.csv")
hourly_calories <- read.csv("hourlyCalories_merged.csv")
hourly_intensities <- read.csv("hourlyIntensities_merged.csv")
hourly_steps <- read.csv("hourlySteps_merged.csv")
weight_info <- read.csv("weightLogInfo_merged.csv")

```

# Data cleaning

rename columns and convert data types to be compatible

```{r}
(hourly_calories <- rename(hourly_calories, hourly_Calories = Calories )) 
(hourly_intensities <-
    rename(hourly_intensities, hourly_TotalIntensity = TotalIntensity, hourly_AverageIntensity = AverageIntensity))

daily_activity$Id <- as.character(daily_activity$Id)
daily_activity$date <- as.Date(daily_activity$ActivityDate, format = "%m/%d/%Y")
str(daily_activity)
head(daily_activity$ActivityDate)


sleepday$Id <- as.character(sleepday$Id)
sleepday$date <- as.Date(sleepday$SleepDay, format = "%m/%d/%Y")
str(sleepday)
head(sleepday)
         

hourly_calories$Id <- as.character(hourly_calories$Id)
hourly_calories$date <- 
  as.POSIXct(hourly_calories$ActivityHour, format = "%m/%d/%Y %I:%M:%S %p" )
str(hourly_calories)
head(hourly_calories)


hourly_intensities$Id <- as.character(hourly_intensities$Id)
hourly_intensities$date=as.POSIXct(hourly_intensities$ActivityHour,
                                             format="%m/%d/%Y %I:%M:%S %p")
str(hourly_intensities)
head(hourly_intensities)


hourly_steps$Id <- as.character(hourly_steps$Id)
hourly_steps$date = as.POSIXct(hourly_steps$ActivityHour, 
                                       format = "%m/%d/%Y %I:%M:%S %p")
str(hourly_steps)
head(hourly_steps)



weight_info$Id <- as.character(weight_info$Id)
weight_info$date <- as.Date(weight_info$Date, format = "%m/%d/%Y" )
str(weight_info)
head(weight_info)

```

check duplicates and any missing values 

```{r}
sum(duplicated(daily_activity))
sum(duplicated(sleepday))
sum(duplicated(hourly_calories))
sum(duplicated(hourly_intensities))
sum(duplicated(hourly_intensities))
sum(duplicated(weight_info))
```

there are three duplicates in sleepday let's remove its.

```{r}
sleepday <- unique(sleepday)
str(sleepday)

```

```{r}
any(is.na(daily_activity))
any(is.na(hourly_calories))
any(is.na(sleepday))
any(is.na(hourly_steps))
any(is.na(hourly_intensities))
any(is.na(weight_info))
sum(is.na(weight_info))
```

all missing data in the fat column so let's remove it 

```{r}
weight_info <- weight_info[, !colnames(weight_info) == "Fat"]
colnames(weight_info)
```

let's summarise data to know where we reached and how to deal with these data.

```{r}
summary(daily_activity)
```

daily_activity dataset have data are not compatible because the mean of veryActiveMinutes column
smaller than lightlyActiveMinutes column and that doesn't make sense. 

```{r}
summary(sleepday)
summary(hourly_calories)
summary(hourly_intensities)
summary(hourly_steps)
summary(weight_info)
```

# Data analysis

create New column is name days to add days of week.

```{r}
daily_activity <- daily_activity %>% mutate(days = weekdays(date))  
colnames(daily_activity)
str(daily_activity)
```

create New column to separate hours from date in that new column.

```{r}
hourly_steps <- hourly_steps %>% mutate(days = weekdays(date)) %>% 
  mutate(hours = format(date, format = "%H:%M:%M %p")) %>% 
  mutate(hours = recode(hours, "00:00:00 AM" = "12:00:00 AM"))
colnames(hourly_steps)
str(hourly_steps)

hourly_calories <- hourly_calories %>% mutate(days = weekdays(date)) %>% 
  mutate(hours = format(date, format = "%H:%M:%M %p")) %>% 
  mutate(hours = recode(hours, "00:00:00 AM" = "12:00:00 AM"))
colnames(hourly_calories)
str(hourly_calories)

hourly_intensities <- hourly_intensities %>% mutate(days = weekdays(date)) %>% 
  mutate(hours = format(date, format = "%H:%M:%M %p")) %>% 
  mutate(hours = recode(hours, "00:00:00 AM" = "12:00:00 AM"))
colnames(hourly_intensities)
str(hourly_intensities)
```

Let's classify the values into conventional values.

```{r}
daily_activity$activity_level <- ifelse(daily_activity$TotalSteps < 5000, "Sedentary",
                                        ifelse(daily_activity$TotalSteps < 7499, "Low Active",
                                               ifelse(daily_activity$TotalSteps < 9999, "moderate Active",
                                                      ifelse(daily_activity$TotalSteps < 12500, "Active", 
                                                             "Highly Active"))))


daily_activity$Calories_burned <- ifelse(daily_activity$Calories < 1900, "very low calories burned",
                                  ifelse(daily_activity$Calories < 2200, "low calories burned",
                                         ifelse(daily_activity$Calories < 2500, "moderate calories burned",
                                                ifelse(daily_activity$Calories < 2900, "high calories burned",
                                                       "very high calories burned"))))


weight_info$Weigh_level <- ifelse(weight_info$WeightKg < 55, "thin",
                                  ifelse(weight_info$WeightKg < 65, "under weight",
                                         ifelse(weight_info$WeightKg < 85 , "Normal",
                                                ifelse(weight_info$WeightKg < 100 , "over weight", "Fat"))))

sleepday$TotalHoursAsleep <- sleepday$TotalMinutesAsleep /60
sleepday$sleep_duration <- ifelse(sleepday$TotalHoursAsleep < 5 , "very low sleep duration",
                                  ifelse(sleepday$TotalHoursAsleep < 7, "low sleep duration",
                                         ifelse(sleepday$TotalHoursAsleep <= 8,"normal sleep duration",
                                                ifelse(sleepday$TotalHoursAsleep < 10, "Long sleep duration",
                                                       "very long sleep duration")))) 


```

let's merge data to do visualization.

```{r}
daily_bellabeat_data <- inner_join(sleepday, daily_activity, by=c('Id', 'date'))
daily_bellabeat_data_weigt <- inner_join(daily_bellabeat_data,weight_info,by=c('Id','date'))
str(daily_bellabeat_data)
str(daily_bellabeat_data_weigt)

hourly_bellabeat_data <- inner_join(hourly_steps, hourly_intensities,
                                    by = c('Id', 'date', 'hours', 'days', 'ActivityHour'))
hourly_bellabeat_data <- inner_join(hourly_bellabeat_data, hourly_calories,
                                    by = c('Id', 'date', 'hours', 'days', 'ActivityHour'))
str(hourly_bellabeat_data)
```

# Data Visualization

## the plots in r plots file have a good view than that in Markdown

visualize Count Activity Level

```{r}
ggplot(data = daily_bellabeat_data, aes(x= activity_level, fill = activity_level)) +
  geom_bar() + labs(title = "Count Activity level")
```

visualize Count Calories Burned

```{r}
ggplot(data = daily_bellabeat_data, aes(x= Calories_burned, fill = Calories_burned)) +
  geom_bar() + labs(title = "Count Calories Burned")
```

Visualize Count Weight Level

```{r}
ggplot(data = weight_info, aes(x= Weigh_level, fill = Weigh_level)) + 
  geom_bar() + labs(title = "Count Weight Level")
```

Visualize Count Sleep duration 

```{r}
ggplot(data = sleepday, aes(x = sleep_duration, fill = sleep_duration)) + 
  geom_bar() + labs(title = "Count Sleep duration")
```

visualize Activity Level through days

```{r}
daily_bellabeat_data$days <- factor(daily_bellabeat_data$days, 
                                    levels = c("Sunday", "Monday", "Tuesday", 
                                               "Wednesday", "Thursday", "Friday", "Saturday"))

ggplot(data = daily_bellabeat_data, aes(x = days, y = activity_level, fill = activity_level)) +
  geom_col() +
  labs(title = "Activity Level through Days")
```

Visualize Sleep duration through days 

```{r}
ggplot(data = daily_bellabeat_data, aes(x = days, y = sleep_duration, fill = sleep_duration)) +
  geom_col() + 
  labs(title = "Sleep duration through Days")
```

Visualize weight level vs calories burned 

```{r}
daily_bellabeat_data_weigt$Calories_burned <- factor(daily_bellabeat_data_weigt$Calories_burned,
                                                     levels = c("very low calories burned","low calories burned",
                                                                "moderate calories burned", "high calories burned",
                                                                "very high calories burned" ))


daily_bellabeat_data_weigt$Weigh_level <- factor(daily_bellabeat_data_weigt$Weigh_level,
                                                 levels = c("thin", "under weight",
                                                            "Normal", "over weight", "Fat" ))

ggplot(data = daily_bellabeat_data_weigt, aes(x = Weigh_level, y= Calories_burned, color = Id)) +
  geom_point() + 
  labs(title = "weight Level vs Calories burned ")
```

visualize hourly Active intensity per hours

```{r}
hourly_bellabeat_intensity <- hourly_bellabeat_data %>% 
  group_by(hours) %>% 
  summarise(hourly_Active_intensity = mean(hourly_TotalIntensity))
ggplot(data = hourly_bellabeat_intensity, aes(x= hours, y= hourly_Active_intensity, fill = hourly_Active_intensity)) + 
         geom_point(stat = "identity") + theme(axis.text.x = element_text(angle = 90))
```

visualize hourly Calories Burned per hours

```{r}
hourly_bellabeat_calories <- hourly_bellabeat_data %>% 
  group_by(hours) %>% 
  summarise(hourly_calories_Burned = mean(hourly_Calories)) 

ggplot(data = hourly_bellabeat_calories, aes(x= hours, y= hourly_calories_Burned, fill = hourly_calories_Burned)) +
  geom_point(stat = "identity") +  theme(axis.text.x = element_text(angle = 90))
  
```

visualize hourly total steps per hours

```{r}
hourly_bellabeat_totalSteps <- hourly_bellabeat_data %>%
  group_by(hours) %>% 
  summarise(hourly_TotalSteps = mean(StepTotal))

ggplot(data = hourly_bellabeat_totalSteps, aes(x= hours, y= hourly_TotalSteps, fill = hourly_TotalSteps)) +
  geom_point(stat = "identity") + theme(axis.text.x = element_text(angle = 90))
```

The data don't target a certain class of people like sports people and overweight people because 
the majority of usage these accassories are under weight people and there are not active intensity or high calories burned and not even hight steps in the morning based on plots depand on hours that
means no sport is being practiced and fat people have low calories burned and fat people don't see these accassories attractive, and majority of people have low sleep duration .

I see that bellabeat should care about sports people by add options to them and also over weight people and There should be messages to keep users healthy

